Deep Learning

### 2015-04-23 deepreview.tex
put back the TDNN citation
Geoff's new version with couple of typos fixed

% deep learning: learning representations
Deep learning is a set of methods that allows a machine to be fed with
raw data and to automatically discover the representations needed for
detection or classification. The name comes from the fact that deep learning machines are
multilayer stacks of trainable modules, each of which turns the
representation of the data on its input into a more abstract and
higher-level representation on its output. The output representation
amplifies aspects of the input that are important for discrimination
and suppresses irrelevant variations. An image, for example, comes in the form of an array of
pixel values and the learned features in the first layer of representation
typically represent the presence
or absence of edges at  particular orientations and locations in
the image. The second layer typically detects motifs by spotting particular
arrangements of edges, regardless of small variations in the edge positions
The third layer  may assemble motifs into larger
combinations that correspond to parts of familiar
objects, and subsequent layers would detect objects as combinations of
these parts. The key aspect of deep learning is that these layers of
features are not designed by human engineers: {\em They are learned from
data using a general purpose learning procedure.}

% preview of deep learning results
Deep learning is making major advances in solving problems that have
resisted the best attempts of the Artificial Intelligence community
for many years. It has turned out to be very good at discovering
intricate structures in large, high-dimensional data and is therefore
applicable to many domains of science, business, and government. In
addition to beating records in image
recognition~\citep{Krizhevsky-2012-small,farabet-pami-13,tompson-nips-14,szegedy-2014}
and speech
recognition~\citep{Hinton-et-al-2012,Sainath-et-al-ICASSP2013}, it has
beaten other ML techniques at predicting the activity of potential
drug molecules~\citep{Ma-et-al-2015}, detecting new fundamental
particles~\citep{Ciodaro-et-al-2012,Melis-Higgs-boson-competition-2014},
% YB thinks that Ciodaro-et-al might be {YANNparticles} 
% classifying galaxies, ~\citep{YANNgalaxies}, % YB while waiting for bib entry, 
reconstructing brain circuits~\citep{helmstaedter-nature-2013}, and
predicting the effects of mutations in non-coding DNA on gene
expression and disease~\citep{Keung-et-al-2014,Xiong-et-al-Frey-science2015}.

Perhaps more surprisingly, deep learning has produced extremely
promising results for a variety of tasks in natural language
understanding~\citep{collobert:2011b}, particularly topic
classification, %~\citep{topic-classification},  % YB while waiting for bib entry
sentiment
analysis~\citep{Glorot+al-ICML-2011-small}, 
question-answering~\cite{Bordes-et-al-EMNLP2014}, and language
translation~\cite{Jean-et-al-arxiv2014,Sutskever-et-al-NIPS2014}.

@@

% shallow linear classifiers
Many practical applications of ML use linear classifiers on top of
hand-engineered features. A two-class linear classifier computes a
weighted sum of the feature vector components. If the weighted sum is
above a threshold, the input is classified as belonging to a
particular category. % Multiple classes can be handled by using multiple
% two-class classifiers, each classifying one class against all
% others. 
% The learning procedure adjusts the weights of the linear
% combination, for which the gradient is very easy to compute. Linear
% classifiers have been widely used for many decades, starting with the
% Perceptron model~\citep{Rosenblatt57} and the still-popular logistic
% regression~\cite{Hastie2001}.
% , and the linear support vector
% machine~\cite{Vapnik+Chervonenkis-1964}. %{Boser92}.

% features: solving the selectivity-invariance dilemma
It has been known since the 1960s that linear classifiers can only
carve their input space into very simple regions, namely
half-spaces separated by a hyperplane~\citep{Duda-Hart}.  But problems
such as image and speech recognition require the input-output function
to be insensitive to irrelevant variations of the input, such as
variations in position, orientation or illumination of an object, or
variations in the pitch or accent of speech, while being very sensitive to
particular minute variations, such as the difference between a white
wolf and a samoyed (a breed of wolf-like white dog).  At the pixel
level, images of two samoyeds in different poses and in different
environments may be very different from each other, while two images
of a samoyed and a wolf in the same position and on similar
backgrounds may be very similar to each other. A linear classifier,
or any other ``shallow'' classifier operating on raw pixels could not
possibly distinguish the latter two, while putting the former two in
the same category. This is why shallow classifiers require a good
feature extractor that solves the {\em selectivity-invariance dilemma},
{\it i.e.} that produces representations that are {\em selective} to the
aspects of the image that are important for discrimination while being {\em invariant} to 
irrelevant aspects like the pose of the animal.
%
% classical extensions of simple classifiers
% YB: I propose to mostly remove the paragraph below (but keep the Scholkopf book ref)
% There have been numerous attempts, going back to the early 1960s, to
% increase the flexibility of classifiers so as to allow them to
% separate categories from one another by more complex surfaces than
% flat hyperplanes. A popular example is the class of kernel methods~\citep{Scholkopf+Smola-2002book}, % added ref for BS
%particularly the Support Vector Machine~\citep{Boser92} which
% represents inputs by a vector whose components are similarities to
% each of the training samples computed through a so-called {\em kernel
%   function}. While these methods are good general-purpose classifiers,
% they do not perform well for tasks such as image or speech recognition
% in which the input space is very high-dimensional and the input-output
% mapping is so complex that computing similarities using a simple
% kernel function is ineffective. 
Designing good feature extractors traditionally required considerable amounts
of engineering skills and domain expertise. But this can all be avoided if
good features can be learned automatically using a general purpose
learning procedure.  That is the key advantage of deep
learning.

@@

% backprop
How do we train multilayer architectures?

From the earliest days of pattern
recognition~\citep{selfridge,Rosenblatt57} researchers have aimed to
replace hand-engineered features by trainable multilayer networks,
but despite its extreme simplicity, the solution was not widely
understood until
the mid 1980s and was largely forsaken by the ML community until
recently. As it turns out, training multilayer architectures can be
done by simple stochastic gradient descent. As long as the modules are
relatively smooth functions of their inputs and of their internal
weights, one can compute gradients using the {\em backpropagation
  procedure}.

The idea that this could be done, and that it worked, was discovered independently by
several different groups in the 1970s and
1980s~\citep{Werbos74,Parker85,LeCun85,RHW}. It spurred a wave of interest for
neural networks, becoming quite popular in the late 1980s, but falling out of favor
between the mid 1990s until their renaissance in the late 2000s.

The backpropagtion procedure to compute gradient of an objective
function with respect to the weights of a multilayer stack of modules
is nothing more than a practical application of the chain rule for
derivatives. The key insight is that the derivative (or gradient) of
the objective with respect to the {\it input} of a module can be
computed by working backwards from the gradient with respect to the
{\em output} of that module (or the input of the subsequent
module), as illustrated in Figure~\ref{fig:backprop-box}.
% Basically, take a module $M$ inside the network, whose input
%is $x$, weight vector is $w$, and output is $y = M(x,w)$. If one knows
%the gradient $\deri{E}{y}$ of the objective function $E$ with respect
%to the module's output $y$, one can blindly apply chain rule to
%compute the gradient with respect to the input:
%\[
%  \deri{E}{x} = \deri{y}{x}\deri{E}{y}
%\]
%This equation works even when $x$ and $y$ are vectors, in which case
%$\deri{E}{x}$ and $\deri{E}{y}$ are gradient vectors, and
%$\deri{y}{x}$ is a {\em Jacobian matrix} of $M$ where element $(i,j)$
%indicates by how much output $j$ would increase proportionally if
%input $i$ were increased by a tiny amount $\Delta y_j =
%\left(\deri{y}{x}\right)_{ij} \Delta x_i$. 
The backpropagation equation can be applied repeatedly to propagate
gradients through all modules, starting from the output at the top 
all the way to the bottom. Once these
gradients have been computed, it is straightforward to compute the gradients 
with respect to the weights of each module. 
%Figure~\ref{fig:backprop-box})
%shows how this works out when the network is composed of layers of
%weight matrices interspersed with scalar non-linearities.
%% the notation in the figure should be updated.

% deep neural nets
Many applications of deep learning use feedforward neural network
architectures, illustrated in Figure~\ref{fig:backprop-box}, which
learn to map a fixed-size input ({\it e.g.} an image) to a fixed-size
output ({\it e.g.} a probability for each one of several categories). To go
from one layer to the next, a set of units compute a weighted
sum of their inputs from the previous layer and pass the result
through a non-linear function. The most common non-linear function is
the so-called {\em ReLU} (Rectified Linear Unit) which is simply the
half-wave rectifier $f(z) = \max(z,0)$. In past decades, neural nets
used smoother non-linearities, such as $\tanh(z)$ or $1/(1+\exp(-z))$ but the ReLU 
typically learns much faster in networks with many layers. Units that are not in the
output layer are traditionally called {\em hidden units}. The early
layers can be seen as distorting the input in a non-linear way so that
categories become linearly separable by the last layer, as indicated
in figure~\ref{fig:backprop-box}.

@@

\section{Convolutional networks}

One of the most widely used deep learning architectures is the
convolutional neural network, or
ConvNet~\cite{lecun-90c,lecun-98}. ConvNets are designed to process
data that come in the form of multiple arrays, for example a color
image composed of three 2D arrays containing pixel intensities in the
three color channels. Many data modalities are in the form of multiple
arrays: 1D for signals and sequences including language, 2D for images
or audio spectrograms, 3D for video or volumetric images.  There are
four key ideas behind ConvNets that take advantage of the properties
of natural signals: local connections, shared weights,
pooling, and the use of many layers.

@@

The convolutional and pooling layers in ConvNets are directly inspired
by the classic notions of simple cells and complex cells in visual
neuroscience~\citep{Hubel62}, and the overall architecture is
reminiscent of the LGN-V1-V2-V4-IT hierarchy in the visual cortex 
ventral pathway~\citep{Felleman+VanEssen-1991}. When ConvNet models and
monkeys are shown the same picture, the activations of high-level
units in the ConvNet explains half the variance of random sets of 160
neurons in the monkey's
inferro-temporal cortex~\citep{cadieu-plos-2014}. ConvNets have their roots
in Fukushima's Neocognitron~\citep{fukushima-82} whose architecture
was somewhat similar but did not have an end-to-end supervised
learning algorithm such as backpropagation. 
% A primitive one-dimensional convnet called a time-delay
% neural net was used for the recognition of phonemes and simple 
% words~\cite{WaibelIEEE,bottou-89}

There have been numerous applications of convolutional networks going
back to the early 1990s, starting with time-delay neural networks for
speech recognition and document reading~\citep{lecun-98}. 
The document reading system used a ConvNet trained jointly with
a probabilistic model that implemented language constraints. By the
late 1990s this system was reading over 10\% of all the checks in the
US. A number of ConvNet-based OCR and handwriting recognition systems
were later deployed by
Microsoft~\citep{simard-03}.
% including for Arabic and
% Chinese~\citep{abdulkader-iwfhr-06,chellapilla-iwfhr-06a}. 
ConvNets
were also experimented with in the early 1990s for object detection 
in natural images, including faces and
hands~\cite{vaillant-monrocq-lecun-94,nowlan-platt-95} and for face
recognition~\cite{lawrence-tnn-1997}.

@@

%representations allow a lot of parallel computation at the feature
%level so that complex tasks can be solved in a very small number of
%sequential steps.
Deep learning theory shows that deep nets can benefit from
an exponential advantage over classical learning algorithm not based on distributed 
representations~\citep{Bengio-localfailure-NIPS-2006-small},
 in two ways, both arising from the power of {\em composition}, so long as the
underlying data generating distribution has such a componential structure, i.e.,
is generated by a hierarchy of composition of factors~\citep{Bengio-Courville-Vincent-TPAMI2013}.
First, learning distributed representations enables generalization to new
combinations of the learned features (e.g. $2^n$ combinations, with $n$ features)
~\citep{Bengio-2009-book,Montufar+Morton-2014}. 
Second composing layers in a deep net brings the potential for another exponential
advantage~\citep{Montufar-et-al-NIPS2014} (exponential in the depth).

@@

Thanks to advances in their architecture~\citep{Hochreiter+Schmidhuber-1997,ElHihi+Bengio-nips8-small} 
and ways of training them~\citep{Sutskever-thesis2012,Pascanu+al-ICML2013-small}, RNNs
have been found to be very good at predicting the next character in
text~\citep{Sutskever-et-al-ICML2011} or the next word in a
sequence~\citep{Mikolov-et-al-NIPS2013} but they can also be used for
more complex tasks.  After reading an English sentence one word at a
time, for example, an English ``encoder'' network can be trained so
that the final state vector of its hidden units is a good
representation of the thought expressed by the sentence.  This thought
vector can then be used as the initial hidden state of (or as extra
input to) a jointly trained French ``decoder'' network which outputs a
probability distribution for the first word of the French
translation. If a particular first word is chosen from this
distribution and provided as input to the decoder network it will then
output a probability distribution for the second word of the
translation and so on until a full stop is
chosen~\citep{Bahdanau-et-al-ICLR2015-small,Sutskever-et-al-NIPS2014}.
Overall, this process generates sequences of French words according to
a probability distribution that depends on the English sentence.

This rather naive way of performing machine translation has quickly become
competitive with the state-of-the-art and this raises serious doubts about
whether understanding a sentence requires anything like internal symbolic
expressions that are manipulated by using inference rules.
%%binding their symbols to variables
%%in judiciously chosen rules of inference, as was previously thought by
%%many early AI researchers. It is 
%%The encoder-decoder translation
%%networks do not work by using internal rules of inference in which
%%variables get bound to symbols. They just use big state vectors, big weight
%%matrices and scalar non-linearities to get the job done. 
%%Deriving conclusions using only the form of the premises (which is
%%what formal logic does) is a powerful
%%technique that people can eventually master, but the translation
%%results with RNNs suggest that it may not be a good model of
%%the way we normally reason or understand language.
%%RNNs that represent thoughts as big vectors are 
It is more compatible with
the view that everyday reasoning involves many simultaneous analogies that
each contribute plausibility to a 
conclusion~\citep{Lakoff+Johnson-2008,Rogers+McClelland-book2004}.

@@

Beyond simple memorization, Neural Turing Machines and Memory Networks
are being used for task that would normally require to manipulate
symbols and to reason.
Neural Turing Machines can be taught ``algorithms''. Among other
things, they can learn to ouput a sorted list of symbols when its
input consists of an unsorted sequence in which each symbol is
accompanied by a real value that indicates its priority in the list
~\citep{Graves-et-al-arxiv2014}.
% To achieve this it must learn a sorting program purely from examples
% of the input and output sequences
Memory Networks can be trained to keep track of the state of the world
in a setting similar to a text adventure game. After reading a story,
they can answer questions that require complex 
inference~\cite{weston-babi-arxiv-2015}.
% Following a sequence
%of actions in a virtual world described in words, in a scenario
%similar to a text adventure game, the network is shown an
%automatically generate question and taught to produce the
%corresponding answer (also generated automatically).  
In one test
example, the network is shown a 15-sentence version of the Lord of The
Rings and correctly answers questions like ``where is Frodo now?''.
~\citep{weston-memorynet-2014}.

@@

\section{Historical perspective on backpropagation}

In the late 1990s, neural nets and backpropagation were largely
forsaken by the machine learning community and ignored by the computer
vision and speech recognition communities.  It was widely believed
that learning multi-stage feature extractors with little prior knowledge 
was neither useful nor effective. In particular, it was commonly
believed that simple gradient descent procedure get trapped in poor
local minima -- weight configurations for which no small change would
reduce the average error-- despite considerable empirical evidence to
the contrary in the case of ConvNets~\citep{lecun-98}. We now know
that the main reasons for the limited success of backpropagation in
deep networks were that the labeled datasets were typically much too
small and the computers were much too slow. Additional reasons were
that researchers were not initializing the weights properly and were
not using the type of non-linearlity that learns most
rapidly~\citep{lecun-98b,GlorotAISTATS2010-small}.

Poor local minima are rarely a problem with large networks. Regardless
of the initial conditions, the system nearly always reaches solutions
of very similar quality.  Recent results strongly suggest the
landscape is packed with a combinatorially large number of {\em saddle
  points} where the gradient is zero, and the surface curves up in
some dimensions and curves down in
others~\citep{Dauphin-et-al-NIPS2014-small,Choromanska-et-al-AISTATS2015}.
The analysis seems to show that local minima (and saddle points with
only a few downward curving directions) are present in very large
numbers, but almost all of them have very similar values of the
objective function.  Hence it doesn't seem to matter which local
minimum the algorithm finds.
%% Avoiding being trapped near saddle points seems more important.

The first major application of this pre-training approach was in
speech recognition and it was made possible by the advent of fast
Graphics Processing Units (GPUs) that were convenient to
program~\citep{RainaICML09-small} and allowed researchers to train
networks 10 or 20 times faster.  In 2009, the approach was used to map
short temporal windows of coefficients extracted from a soundwave to a
set of probabilities for the various fragments of speech that might be
represented by the frame in the center of the window.  It achieved
record-breaking results on a standard speech recognition benchmark
that used a small vocabulary~\citep{TIMITpaper} and was quickly
developed to give record-breaking results on a large vocabulary
task~\citep{Dahl2012}.  By 2012, versions of the deep net from 2009
were being developed by many of the major speech
groups~\citep{Hinton-et-al-2012} and were already being deployed in
Android phones.  For smaller datasets, unsupervised pre-training helps
to prevent overfitting~\citep{Bengio-Courville-Vincent-TPAMI2013},
leading to significantly better generalization when the number of
labeled examples is small, or in a transfer setting where we have lots
of examples for some ``source'' tasks but very few for some ``target''
task. In this context, unsupervised representation learning helped to
win several transfer learning
competitions~\citep{Bengio-Courville-Vincent-TPAMI2013}.  Once deep
learning had been rehabilitated, it turned out that large labeled
datasets did not need the pre-training stage. Even better results,
especially for very deep networks, could be obtained by using
rectifying non-linearities rather than sigmoid or tanh non-linearities
~\citep{jarrett-iccv-09,Nair-2010-small,Glorot+al-AI-2011-small,Krizhevsky-2012-small,Dahl-et-al-ICASSP2013}. Deep
neural nets are now the dominant method for acoustic modeling in
speech, and ConvNet are dominant for image and video recognition.

>>>>

% deep learning: learning representations
Deep learning is a set of methods that allows a machine to be fed with
raw data and to automatically discover the representations needed for
detection or classification. The name comes from the fact that deep learning machines are
multilayer stacks of trainable modules, each of which turns the
representation of the data on its input into a more abstract and
higher-level representation on its output. The output representation
amplifies aspects of the input that are important for discrimination
and suppresses irrelevant variations. An image, for example, comes in the form of an array of
pixel values and the learned features in the first layer of representation
typically represent the presence
or absence of edges at  particular orientations and locations in
the image. The second layer typically detects motifs by spotting particular
arrangements of edges, regardless of small variations in the edge positions
The third layer  may assemble motifs into larger
combinations that correspond to parts of familiar
objects, and subsequent layers would detect objects as combinations of
these parts. The key aspect of deep learning is that these layers of
features are not designed by human engineers: They are learned from
data using a general purpose learning procedure.

% preview of deep learning results
Deep learning is making major advances in solving problems that have
resisted the best attempts of the Artificial Intelligence community
for many years. It has turned out to be very good at discovering
intricate structures in high-dimensional data and is therefore
applicable to many domains of science, business, and government. In
addition to beating records in image
recognition~\citep{Krizhevsky-2012-small,farabet-pami-13,tompson-nips-14,szegedy-2014}
and speech
recognition~\citep{Hinton-et-al-2012,Sainath-et-al-ICASSP2013}, it has
beaten other ML techniques at predicting the activity of potential
drug molecules~\citep{Ma-et-al-2015}, detecting new fundamental
particles~\citep{Ciodaro-et-al-2012,Melis-Higgs-boson-competition-2014},
% YB thinks that Ciodaro-et-al might be {YANNparticles} 
% classifying galaxies, ~\citep{YANNgalaxies}, % YB while waiting for bib entry, 
reconstructing brain circuits~\citep{helmstaedter-nature-2013}, and
predicting the effects of mutations in non-coding DNA on gene
expression and disease~\citep{Keung-et-al-2014,Xiong-et-al-Frey-science2015}.
Perhaps more surprisingly, deep learning has produced extremely
promising results for a variety of tasks in natural language
understanding~\citep{collobert:2011b}, particularly topic
classification, %~\citep{topic-classification},  % YB while waiting for bib entry
sentiment
analysis~\citep{Glorot+al-ICML-2011-small}, 
question-answering~\cite{Bordes-et-al-EMNLP2014}, and language
translation~\cite{Jean-et-al-arxiv2014,Sutskever-et-al-NIPS2014}.

% our conclusions (which Nature insists on having at the end of the intro).
We believe that deep learning will have many more successes in the
near future because it requires very little hand-engineering so it can
easily take advantage of increases in the amount of available
computation and data. New learning algorithms and new architectures
that are currently being developed for deep neural networks will only
accelerate this progress.

\section{Supervised Learning}

@@

% shallow linear classifiers
Many of the current practical applications of ML use linear classifiers on top of
hand-engineered features. A two-class linear classifier computes a
weighted sum of the feature vector components. If the weighted sum is
above a threshold, the input is classified as belonging to a
particular category. % Multiple classes can be handled by using multiple
% two-class classifiers, each classifying one class against all
% others. 
% The learning procedure adjusts the weights of the linear
% combination, for which the gradient is very easy to compute. Linear
% classifiers have been widely used for many decades, starting with the
% Perceptron model~\citep{Rosenblatt57} and the still-popular logistic
% regression~\cite{Hastie2001}.
% , and the linear support vector
% machine~\cite{Vapnik+Chervonenkis-1964}. %{Boser92}.

% features: solving the selectivity-invariance dilemma
It has been known since the 1960s that linear classifiers can only
carve their input space into very simple regions, namely
half-spaces separated by a hyperplane~\citep{Duda-Hart}.  But problems
such as image and speech recognition require the input-output function
to be insensitive to irrelevant variations of the input, such as
variations in position, orientation or illumination of an object, or
variations in the pitch or accent of speech, while being very sensitive to
particular minute variations, such as the difference between a white
wolf and a samoyed (a breed of wolf-like white dog).  At the pixel
level, images of two samoyeds in different poses and in different
environments may be very different from each other, while two images
of a samoyed and a wolf in the same position and on similar
backgrounds may be very similar to each other. A linear classifier,
or any other ``shallow'' classifier operating on raw pixels could not
possibly distinguish the latter two, while putting the former two in
the same category. This is why shallow classifiers require a good
feature extractor that solves the {\em selectivity-invariance dilemma},
{\it i.e.} that produces representations that are {\em selective} to the
aspects of the image that are important for discrimination while being {\em invariant} to 
irrelevant aspects like the pose of the animal.
%
% classical extensions of simple classifiers
% YB: I propose to mostly remove the paragraph below (but keep the Scholkopf book ref)
% There have been numerous attempts, going back to the early 1960s, to
% increase the flexibility of classifiers so as to allow them to
% separate categories from one another by more complex surfaces than
% flat hyperplanes. A popular example is the class of kernel methods~\citep{Scholkopf+Smola-2002book}, % added ref for BS
%particularly the Support Vector Machine~\citep{Boser92} which
% represents inputs by a vector whose components are similarities to
% each of the training samples computed through a so-called {\em kernel
%   function}. While these methods are good general-purpose classifiers,
% they do not perform well for tasks such as image or speech recognition
% in which the input space is very high-dimensional and the input-output
% mapping is so complex that computing similarities using a simple
% kernel function is ineffective. 
Designing good feature extractors traditionally required a considerable amount
of engineering skill and domain expertise. But this can all be avoided if
good features can be learned automatically using a general purpose
learning procedure.  That is the key advantage of deep
learning.

@@

% backprop
\section{Backpropagation: a simple way to train multilayer architectures}

From the earliest days of pattern
recognition~\citep{selfridge,Rosenblatt57} researchers have aimed to
replace hand-engineered features by trainable multilayer networks,
but despite its simplicity, the solution was not widely
understood until
the mid 1980s.  
% and was largely forsaken by the ML community until recently.
As it turns out, multilayer architectures can be
trained by simple stochastic gradient descent. As long as the modules are
relatively smooth functions of their inputs and of their internal
weights, one can compute gradients using the {\em backpropagation
  procedure}.
The idea that this could be done, and that it worked, was discovered independently by
several different groups in the 1970s and
1980s~\citep{Werbos74,Parker85,LeCun85,RHW}. 
% It spurred a wave of interest for
% neural networks, becoming quite popular in the late 1980s, but falling out of favor
% between the mid 1990s until their renaissance in the late 2000s.

The backpropagation procedure to compute the gradient of an objective
function with respect to the weights of a multilayer stack of modules
is nothing more than a practical application of the chain rule for
derivatives. The key insight is that the derivative (or gradient) of
the objective with respect to the {\it input} of a module can be
computed by working backwards from the gradient with respect to the
{\em output} of that module (or the input of the subsequent
module), as illustrated in Figure~\ref{fig:backprop-box}.
% Basically, take a module $M$ inside the network, whose input
%is $x$, weight vector is $w$, and output is $y = M(x,w)$. If one knows
%the gradient $\deri{E}{y}$ of the objective function $E$ with respect
%to the module's output $y$, one can blindly apply chain rule to
%compute the gradient with respect to the input:
%\[
%  \deri{E}{x} = \deri{y}{x}\deri{E}{y}
%\]
%This equation works even when $x$ and $y$ are vectors, in which case
%$\deri{E}{x}$ and $\deri{E}{y}$ are gradient vectors, and
%$\deri{y}{x}$ is a {\em Jacobian matrix} of $M$ where element $(i,j)$
%indicates by how much output $j$ would increase proportionally if
%input $i$ were increased by a tiny amount $\Delta y_j =
%\left(\deri{y}{x}\right)_{ij} \Delta x_i$. 
The backpropagation equation can be applied repeatedly to propagate
gradients through all modules, starting from the output at the top 
all the way to the bottom. Once these
gradients have been computed, it is straightforward to compute the gradients 
with respect to the weights of each module. 
%Figure~\ref{fig:backprop-box})
%shows how this works out when the network is composed of layers of
%weight matrices interspersed with scalar non-linearities.
%% the notation in the figure should be updated.

% deep neural nets
Many applications of deep learning use feedforward neural network
architectures, illustrated in Figure~\ref{fig:backprop-box}, which
learn to map a fixed-size input ({\it e.g.} an image) to a fixed-size
output ({\it e.g.} a probability for each of several categories). To
go from one layer to the next, a set of units compute a weighted sum
of their inputs from the previous layer and pass the result through a
non-linear function. Currently, the most popular non-linear function
is the rectified linear unit ({\em ReLU}) which is simply
the half-wave rectifier $f(z) = \max(z,0)$. In past decades, neural
nets used smoother non-linearities, such as $\tanh(z)$ or
$1/(1+\exp(-z))$ but the ReLU typically learns much faster in networks
with many layers. Units that are not in the input or output layer are
traditionally called {\em hidden units}. The hidden layers can be seen
as distorting the input in a non-linear way so that categories become
linearly separable by the last layer, as indicated in
figure~\ref{fig:backprop-box}.

@@

% the reduced history section

In the late 1990s, neural nets and backpropagation were largely
forsaken by the machine learning community and ignored by the computer
vision and speech recognition communities.  It was widely believed
that learning useful, multi-stage, feature extractors with little prior knowledge 
was infeasible. In particular, it was commonly
believed that simple gradient descent would get trapped in poor
local minima -- weight configurations for which no small change would
reduce the average error.
%We now know
%that the main reasons for the limited success of backpropagation in
%deep networks with dense connectivity were that the labeled datasets were typically much too
%small and the computers were much too slow. Additional reasons were
%that researchers were not initializing the weights properly and were
%not using the type of non-linearlity that learns most
%rapidly~\citep{lecun-98b,GlorotAISTATS2010-small}.

In practice, poor local minima are rarely a problem with large networks. Regardless
of the initial conditions, the system nearly always reaches solutions
of very similar quality.  Recent results strongly suggest the
landscape is packed with a combinatorially large number of {\em saddle
  points} where the gradient is zero, and the surface curves up in
most dimensions and curves down in the 
remainder~\citep{Dauphin-et-al-NIPS2014-small,Choromanska-et-al-AISTATS2015}.
The analysis seems to show that saddle points with
only a few downward curving directions are present in very large
numbers, but almost all of them have very similar values of the
objective function.  Hence it doesn't much matter which of these
saddle points the algorithm gets stuck at.
%% Avoiding being trapped near saddle points seems more important.

The first major application of this pre-training approach was in
speech recognition and it was made possible by the advent of fast
Graphics Processing Units (GPUs) that were convenient to
program~\citep{RainaICML09-small} and allowed researchers to train
networks 10 or 20 times faster.  In 2009, the approach was used to map
short temporal windows of coefficients extracted from a soundwave to a
set of probabilities for the various fragments of speech that might be
represented by the frame in the center of the window.  It achieved
record-breaking results on a standard speech recognition benchmark
that used a small vocabulary~\citep{TIMITpaper} and was quickly
developed to give record-breaking results on a large vocabulary
task~\citep{Dahl2012}.  By 2012, versions of the deep net from 2009
were being developed by many of the major speech
groups~\citep{Hinton-et-al-2012} and were already being deployed in
Android phones.  For smaller datasets, unsupervised pre-training helps
to prevent overfitting~\citep{Bengio-Courville-Vincent-TPAMI2013},
leading to significantly better generalization when the number of
labeled examples is small, or in a transfer setting where we have lots
of examples for some ``source'' tasks but very few for some ``target''
task.  Once deep learning had been rehabilitated, it turned out the
pre-training stage was only needed for small datasets.

There was, however, one particular type of deep, feed-forward network
that was much easier to train and generalized much better than
networks with full connectivity between adjacent layers. This was the
convolutional neural network~\cite{lecun-90c,lecun-98}. It achieved
many practical successes during the period when neural networks were
out of favor and it has recently been widely adopted by the computer
vision community.

\section{Convolutional neural networks}

Convolutional neural networks (ConvNets) are designed to process
data that come in the form of multiple arrays, for example a color
image composed of three 2D arrays containing pixel intensities in the
three color channels. Many data modalities are in the form of multiple
arrays: 1D for signals and sequences including language, 2D for images
or audio spectrograms, 3D for video or volumetric images.  There are
four key ideas behind ConvNets that take advantage of the properties
of natural signals: local connections, shared weights,
pooling, and the use of many layers.

@@

The convolutional and pooling layers in ConvNets are directly inspired
by the classic notions of simple cells and complex cells in visual
neuroscience~\citep{Hubel62}, and the overall architecture is
reminiscent of the LGN-V1-V2-V4-IT hierarchy in the visual cortex 
ventral pathway~\citep{Felleman+VanEssen-1991}. When ConvNet models and
monkeys are shown the same picture, the activations of high-level
units in the ConvNet explains half the variance of random sets of 160
neurons in the monkey's
inferro-temporal cortex~\citep{cadieu-plos-2014}. ConvNets have their roots
in Fukushima's Neocognitron~\citep{fukushima-82} whose architecture
was somewhat similar but did not have an end-to-end supervised
learning algorithm such as backpropagation. 
A primitive one-dimensional convnet called a time-delay
neural net was used for the recognition of phonemes and simple 
words~\cite{WaibelIEEE,bottou-89}

There have been numerous applications of convolutional networks going
back to the early 1990s, starting with time-delay neural networks for
speech recognition~\citep{TDNN} and document reading~\citep{lecun-98}. 
The document reading system used a ConvNet trained jointly with
a probabilistic model that implemented language constraints. By the
late 1990s this system was reading over 10\% of all the checks in the
US. A number of ConvNet-based OCR and handwriting recognition systems
were later deployed by
Microsoft~\citep{simard-03}.
% including for Arabic and
% Chinese~\citep{abdulkader-iwfhr-06,chellapilla-iwfhr-06a}. 
ConvNets
were also experimented with in the early 1990s for object detection 
in natural images, including faces and
hands~\cite{vaillant-monrocq-lecun-94,nowlan-platt-95} and for face
recognition~\cite{lawrence-tnn-1997}.

@@

%representations allow a lot of parallel computation at the feature
%level so that complex tasks can be solved in a very small number of
%sequential steps.
Deep learning theory shows that deep nets have two different
exponential advantages over classical learning algorithms that do not
use distributed
representations~\citep{Bengio-localfailure-NIPS-2006-small}. Both of
these advantages arise from the power of {\em composition} and depend
on the underlying data generating distribution having an appropriate
componential structure~\citep{Bengio-Courville-Vincent-TPAMI2013}.
First, learning distributed representations enables generalization to
new combinations of the learned features (e.g. $2^n$ combinations,
with $n$ features) ~\citep{Bengio-2009-book,Montufar+Morton-2014}.
Second composing layers in a deep net brings the potential for another
exponential advantage~\citep{Montufar-et-al-NIPS2014} (exponential in
the depth).

@@

Thanks to advances in their architecture~\citep{Hochreiter+Schmidhuber-1997,ElHihi+Bengio-nips8-small} 
and ways of training them~\citep{Sutskever-thesis2012,Pascanu+al-ICML2013-small}, RNNs
have been found to be very good at predicting the next character in
text~\citep{Sutskever-et-al-ICML2011} or the next word in a
sequence~\citep{Mikolov-et-al-NIPS2013} but they can also be used for
more complex tasks.  After reading an English sentence one word at a
time, for example, an English ``encoder'' network can be trained so
that the final state vector of its hidden units is a good
representation of the thought expressed by the sentence.  This thought
vector can then be used as the initial hidden state of (or as extra
input to) a jointly trained French ``decoder'' network which outputs a
probability distribution for the first word of the French
translation. If a particular first word is chosen from this
distribution and provided as input to the decoder network it will then
output a probability distribution for the second word of the
translation and so on until a full stop is
chosen~\citep{Bahdanau-et-al-ICLR2015-small,Sutskever-et-al-NIPS2014}.
Overall, this process generates sequences of French words according to
a probability distribution that depends on the English sentence.
This rather naive way of performing machine translation has quickly become
competitive with the state-of-the-art and this raises serious doubts about
whether understanding a sentence requires anything like internal symbolic
expressions that are manipulated by using inference rules.

%%binding their symbols to variables
%%in judiciously chosen rules of inference, as was previously thought by
%%many early AI researchers. It is 
%%The encoder-decoder translation
%%networks do not work by using internal rules of inference in which
%%variables get bound to symbols. They just use big state vectors, big weight
%%matrices and scalar non-linearities to get the job done. 
%%Deriving conclusions using only the form of the premises (which is
%%what formal logic does) is a powerful
%%technique that people can eventually master, but the translation
%%results with RNNs suggest that it may not be a good model of
%%the way we normally reason or understand language.
%%RNNs that represent thoughts as big vectors are 
It is more compatible with
the view that everyday reasoning involves many simultaneous analogies that
each contribute plausibility to a 
conclusion~\citep{Lakoff+Johnson-2008,Rogers+McClelland-book2004}.

@@

Beyond simple memorization, Neural Turing Machines and Memory Networks
are being used for tasks that would normally require reasoning and symbol manipulation.
Neural Turing Machines can be taught ``algorithms''. Among other
things, they can learn to output a sorted list of symbols when their
input consists of an unsorted sequence in which each symbol is
accompanied by a real value that indicates its priority in the list
~\citep{Graves-et-al-arxiv2014}.
% To achieve this it must learn a sorting program purely from examples
% of the input and output sequences
Memory Networks can be trained to keep track of the state of the world
in a setting similar to a text adventure game and after reading a story,
they can answer questions that require complex 
inference~\cite{weston-babi-arxiv-2015}.
% Following a sequence
%of actions in a virtual world described in words, in a scenario
%similar to a text adventure game, the network is shown an
%automatically generate question and taught to produce the
%corresponding answer (also generated automatically).  
In one test
example, the network is shown a 15-sentence version of the Lord of The
Rings and correctly answers questions like ``where is Frodo now?''.
~\citep{weston-memorynet-2014}.

@@
